---
output:
  md_document:
    variant: gfm
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, cache = FALSE,
  comment = "# ",
  fig.path = "tools/README-",
  dpi = 67
)
```



## Cooperative Penalized Regression (CooPeR) ⚠️ Fork Notice ⚠️

This repository is a fork of the original [`fwelnet` implementation by Tay et al](https://github.com/kjytay/fwelnet).  
The goal of this fork is support a research project based on the extension of `fwelnet` to survival settings and competing event analysis.
For this purpose, some of the internal functions were modified in a way that is not directly compatible with the original implementation, hence this project currently stands on its own, rather than being a pull-request to the original implementation.

This version, renamed to `cooper`, supports targets of the form `Surv(time, event)` from the `survival` package, and fits `fwelnet` through the minimization of the Cox deviance.
In addition, it adds the `cooper()` function for a specific implementation of the multi-task algorithm proposed by Tay et al. which fits cause-specific Cox models.
As of now, the implementation is tailored towards a specific research question rather than general use.

Content from the original repository follows.

# Feature-Weighted Elastic Net (fwelnet)

`fwelnet` is a package that fits the __*feature-weighted elastic net (fwelnet)*__, a variant of the elastic net which has feature-specific penalties. These penalties are based on additional information that the user has on the features. `fwelnet` works with continuous and binary responses. For details, please see the [preprint](https://arxiv.org/abs/2006.01395). For a short tutorial on how to use the package, please see the vignette in the `vignettes/` folder.


## An example

Here is a simple example to illustrate how to use this package. First, let's generate some data. In this example, we assume that we have 40 features, and that these features come in 4 groups of 10. The response is a linear combination of the features from the first 2 groups with additive Gaussian noise.
```{r}
set.seed(1)
n <- 100
p <- 40
groups <- list(1:10, 11:20, 21:30, 31:40)  # which features belong to which group
x <- matrix(rnorm(n * p), nrow = n, ncol = p)
beta <- matrix(rep(1:0, each = 20), ncol = 1)
y <- x %*% beta + rnorm(n)
```

In order to fit a fwelnet model, we have to define a feature information matrix. In our example, we have $\mathbf{Z} \in \mathbb{R}^{40 \times 4}$, with $z_{jk} = 1\{ \text{feature } j \text{ belongs to group } k \}$.
```{r}
# generate Z matrix
z <- matrix(0, nrow = p, ncol = length(groups))
for (i in 1:length(groups)) {
    z[groups[[i]], i] <- 1
}
```

Once `z` is specified, we can fit the fwelnet model with `fwelnet()`.
```{r}
library(fwelnet)
fit <- fwelnet(x, y, z)
```

"fwelnet" objects are equipped with `predict` and `coef` methods which allow the user to make predictions on new data and to view the model coefficients. By default predictions and coefficients are returned for the whole `lambda` path.
```{r}
# predictions for first 5 observations at 20th lambda value
predict(fit, x[1:5, ])[, 20]

# coefficients at the 20th lambda value (including intercept)
as.numeric(coef(fit)[, 20])
```
